---
title: "Total polyphenols"
output: html_notebook
---
```{r}
library(tidyverse)
library(here)
library(car)
```

Data was collected by Amma in Spring 2019.

```{r}
tp <- read_csv(here("data", "2017", "raw", "2017 total polyphenols.csv"))
```

```{r}
tp
```

- Sample: cultivar (all were Qing Xin Da Mao)
- Standard Deviation: from averaging triplicates
- Polyphenols: galic acid equivalents in mg/mL (i think)
- Coefficient of Variation: from averaging triplicates
- Density: leafhopper density at end of experiment in insects/young leaf
- mg tea: mg of tea powder
- mg/g GAE: galic acid equivalents in mg/g of tea powder

```{r}
m <- lm(`mg/g GAE` ~ poly(Density, 2), data = tp%>% filter(`mg/g GAE` > 4))
m2 <- lm(`mg/g GAE` ~ Density, data = tp%>% filter(`mg/g GAE` > 4))
m0 <- lm(`mg/g GAE`~ 1, data = tp%>% filter(`mg/g GAE` > 4))
AIC(m, m2, m0)
summary(m)
Anova(m)
```


```{r}
ggplot(tp %>% filter(`mg/g GAE` > 4), aes(x = Density, y = `mg/g GAE`)) +
  geom_point() +
  geom_smooth(method = "lm", formula = "y ~ poly(x, 2)")
```

Seems like maybe that one low point is an outlier.  Could be worth re-doing assay?
