---
title: "2017 Leafhopper GCMS analysis"
author: "Eric R. Scott"
date: "2019-08-01"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999) #turn off scientific notation
```

*Last compiled: `r Sys.Date()`*

```{r packages, include=FALSE}
library(tidyverse)
library(here)
library(readxl)
#for RDA:
library(vegan)
library(RVAideMemoire)
#for figures:
library(cowplot)
#for step and hinge models
library(chngpt)
```

# Purpose

This is the main analysis of volatile data for the paper. Multivariate analysis with redundancy analysis (RDA).  I decided to go with RDA over PLSR because I like that it gives me the percentage of variation explained by the predictor variables (herbivory proxies) whereas PLSR just gives a R2 for the axis (how much variation in X and Y does the *axis* explain).  I then identify biomarkers as compounds with significant correlations to the first (only) RDA axis, and do some univariate tests on them to see if they best fit a null (intercept only), line, step function, or hinge and then plot the results and produce a table.

# Load Data
Read in cleaned GCMS data (which includes leafhopper density already) and leafhopper damage data (from image analysis)

```{r data, echo=TRUE}
gc_wide <- read_rds(here("data", "cleaned", "2017_gcms_wide.rds"))
gc_tidy <- read_rds(here("data", "cleaned", "2017_gcms_tidy.rds")) %>% rename(compound = Compound)
```

```{r}
annotations.raw <- 
  read_excel(here("data", "cleaned", "Compound annotation updated.xlsx"), na = "NA") %>%
  janitor::clean_names()
```


## Data Dictionary

`gc_tidy`: A tidy dataframe of all the chemistry and other data

- `sample`: Sample name
- `cultivar`: In this year, all plants were Qing Xin Da Mao, but I included this column for uniformity with the 2018 experiment.
- `density_start`: The density of leafhoppers applied at the start of the experiment in **insects/young leaf**
- `density_end`: Ending density in **insects/young leaf** 
- `mean_percent_damage`: Percentage of damaged pixels averaged across all leaves for that sample
- `twister_damage`: Percentage of damaged pixels for the leaf the volatiles were sampled from.
- `No.`: Compound number.  This comes from the Ion Analytics methods file.
- `Compound`: Compound name
- `RPA`: Relative peak area
- `rt`: retention time, in minutes
- `ri`: retention index
- `present`: logical. Was the compound detected in a particular sample?

`gc_wide`: A wide version with columns for each `Compound` with `RPA` as the value

# Load Functions
Custom functions needed for analysis:

```{r}
source(here("R", "functions.R"))
```


# Data pre-treatment
Most data pre-treatment was done as part of creating the .rds files I read in here. I already determined that log transformation followed by scaling is probably a good idea to improve normality.  I also determined that removing one outlier (sample Q8) would be prudent.

```{r}
#columns that aren't compound RPAs
metavars <- c("sample", "cultivar", "density_start", "density_end", "mean_percent_damage", "twister_damage")
```
 


```{r}
gc_wide.logscale <- 
  gc_wide %>% 
  mutate_at(vars(-metavars), log) %>%
  mutate_at(vars(-metavars), scale) %>%
  #strip atrributes left by scale() that interfere with some other functions down the road
  mutate_at(vars(-metavars), rm_attr) %>% 
  #also log-transform leaf damage
  mutate(log.twister = log(twister_damage),
         log.mean_damage = log(mean_percent_damage)) %>% 
  #remove outlier found in wrangling script.
  filter(sample != "Q8")

gc_tidy <-
  gc_tidy %>%
  mutate(log.twister = log(twister_damage),
         log.mean_damage = log(mean_percent_damage)) %>% 
  filter(sample != "Q8")

#update list of columns that aren't RPAs
metavars <- c(metavars, "log.twister", "log.mean_damage")
#set up chemistry data for RDA
chemdata <-
  gc_wide.logscale %>% 
  select(-metavars) %>%
  as.data.frame()

rownames(chemdata) <- gc_wide.logscale$sample
```

# Descriptive statistics
## How many compounds total?

```{r}
length(unique(gc_tidy$compound))
```

## How many compounds are found in all samples?

```{r}
common <-
  gc_tidy %>% 
  group_by(compound) %>% 
  summarize(nsample = sum(present)) %>% 
  filter(nsample > 17) 
nrow(common)
# common
```

## Summary table
Create a summary table and join with annotations, rename unknowns in order by retention index

```{r}
meanRIs <-
  gc_tidy %>% 
  group_by(compound) %>% 
  summarize(meanRI = mean(ri, na.rm = TRUE))

annotations2 <- left_join(meanRIs, annotations.raw, by = "compound")
  
unks <-
  annotations2 %>% 
  filter(is.na(cas)) %>% 
  arrange(meanRI) %>% 
  add_column(unk_name = 1:nrow(.))

annotations <- 
  left_join(annotations2, unks) %>% 
  mutate(pretty_name = ifelse(is.na(unk_name), pretty_name, unk_name)) %>% 
  mutate(pretty_name = ifelse(is.na(pretty_name), compound, pretty_name)) %>% 
  select(compound, cas, pretty_name, meanRI, ri_lab, ri_lit, source, aroma, chemical_family) %>% 
  mutate(chemical_family = snakecase::to_sentence_case(chemical_family))
```



# Analysis
## RDA for density
### Fit RDA model

Only analysis for the regular RDA is shown, because the distance-based RDA gave the same results and you can't get loadings from it.

```{r}
#scaled, transformed
rda_dens <-
  rda(chemdata ~ gc_wide.logscale$density_end)
```

### How much total variance does the experimental design explain?

```{r}
synt_dens <- MVA.synt(rda_dens)
synt_dens[[1]][[2]]
```
10.66%

### Permutation test

Test for significance that this amount of explained variance is higher than the null hypothesis of no effect of the experimental design:

```{r}
anova(rda_dens)
```

Significant effect of ending density on leaf volatiles


## RDA for mean percent damage
### Fit RDA

```{r}
rda_dam <- rda(chemdata ~ gc_wide.logscale$log.mean_damage)
```

### How much total variance does the experimental design explain?

```{r}
synt_dam <- MVA.synt(rda_dam)
synt_dam[[1]][[2]]
```

6.61%

### Permutation Test
Test for significance that this amount of explained variance is higher than the null hypothesis of no effect of the experimental design:

```{r}
anova(rda_dam)
```

Not significant

## RDA for focal DCSE leaf damage
### Fit RDA

```{r}
rda_t <- rda(chemdata ~ gc_wide.logscale$log.twister)
```

### How much total variance does the experimental design explain?

```{r}
synt_t <- MVA.synt(rda_t)
synt_t[[1]][[2]]
```
7.67% 

### Permutation Test
Test for significance that this amount of explained variance is higher than the null hypothesis of no effect of the experimental design:

```{r}
anova(rda_t)
```

Marginally significant.


# Biomarkers

I'm defining biomarkers as compounds that are significantly correlated to the RDA axis (correlation between axis scores and chemistry data).

## leafhopper density biomarkers

### Extract scores and loadings

```{r}
dens_scores <-
  scores(rda_dens, display = "sites", scaling = 0) %>% 
  as_tibble(rownames = "sample")

dens_loads <- scores(rda_dens, display = "species", scaling = 0) %>% 
  as_tibble(rownames = "compound")
```

### Correlation between data and RDA axis

```{r}
dens_corr <-
  map_df(chemdata,
         ~cor.test(.x, dens_scores$RDA1) %>%
           broom::glance(),
         .id = "compound") %>% 
  select(compound, correlation = estimate, corr.p.value = p.value)
```


### Filter

Filter by either significant correlation to RDA or significant univariate regression. Arrange by absolute value of correlation coefficient.

```{r}
dens_biomarkers <-
  dens_corr %>% 
  filter(corr.p.value <= 0.05) %>% 
  arrange(desc(abs(correlation)))
```

### Univariate regressions

Log transformation improved normality of the whole dataset, so I should be consistent and use that here as well. Log transformation is better than log-link glm in this scenario, because we *want* the variance to be transformed.

```{r}
#Only need to do this on compounds in biomarkers list
dens_biom_data <-
  gc_tidy %>% 
  mutate(log.RPA = log(RPA)) %>% 
  filter(compound %in% dens_biomarkers$compound)
```

I'm going to fit a null (intercept only), linear, step, and hinge model to every compound and choose the winner with AIC.

Null is the simplest (df = 0), folowed by line (df = 1), and step and hinge (df = 2).  I also want to ignore the step and hinge models if the confidence interval for the changepoint overlaps the ends of the x-axis.  That indicates to me that a line or intercept only is a better fit.

```{r}
dens_univar <-
  dens_biom_data %>% 
  group_by(compound) %>% 
  nest() %>% 
  #fit null, linear, step, and hing models
  mutate(null = map(data, ~lm(log.RPA ~ 1, data = .x)),
         line = map(data, ~lm(log.RPA ~ density_end, data = .x)),
         step = map(data, ~chngptm(log.RPA ~ 1, ~ density_end,
                                   family = "gaussian", type = "step", data = .x,
                                   var.type="bootstrap", ci.bootstrap.size=100)),
         hinge = map(data, ~chngptm(log.RPA ~ 1, ~ density_end, 
                                    family = "gaussian", type = "hinge", data = .x))) %>% 
  gather(-compound, -data, key = type, value = model) %>% 
  #extract winning glm from step and hing models
  mutate(fit = ifelse(type %in% c("step", "hinge"), map(model, ~.x$best.fit), model)) %>% 
  arrange(compound) %>% 
  mutate(AIC = map(fit, ~AIC(.x))) %>%
  unnest(AIC)
```

Get model info and coeficients and choose the best model by AIC, penalizing any hinge or step models that have threshold CIs that overlap zero.

```{r}
# This is hacky, but works.  Ran into error when trying to do it all in one dataset.
dens_univar2 <- 
  dens_univar %>% 
  filter(type %in% c("step", "hinge")) %>% 
  mutate(coefs = purrr::map(model, tidy.chngpt)) %>% 
  unnest(coefs) %>% 
  select(-data, -model, -fit) %>% 
  #joins to null and line models
  left_join(dens_univar, .) %>%
  #set AIC to very high if the threshold CI overlaps zero
  mutate(AIC = if_else(chngpt.lower.CI > 0 | is.na(chngpt.lower.CI), AIC, Inf)) %>%
  mutate(AIC = if_else(chngpt.upper.CI < max(gc_tidy$density_end) | is.na(chngpt.upper.CI),
                       AIC, Inf)) %>% 
  #set AIC to very high if coefficient isn't significantly different from zero
  mutate(AIC = if_else(coef.p.value > 0.05, Inf, AIC,missing = AIC)) %>% 
  group_by(compound) %>%
  # choose the best AIC as described above
  filter(AIC == bestAIC(AIC))
```

Create a column that is the slope when linear, the slope after the threshold when hinge, and the difference between intercepts when step.  The idea is to give the reader some idea of the direction and magnitude of the relationship without plotting all 30.

```{r}
dens_univar_out <-
  dens_univar2 %>% 
  mutate(estimate = case_when(type == "step"  ~ `I(density_end>chngpt)`,
                              type == "hinge" ~ `(density_end-chngpt)+`,
                              type == "line"  ~ map_dbl(fit, ~coef(.x)[2]))) %>% 
  select(compound, type, estimate, threshold = chngpt)
```


### Merge to create tables


Biomarker table should have: 

- Compound (pretty name)
- CAS
- RDA loading
- Correlation (with *** for p-values)
- Response type (null, linear, hinge, step)
- Response estimate (linear = slope, hinge = slope after threshold, step = intercept2 - intercept1)
- Change point
- Chemical family (???)
- Aroma

```{r}
table1 <-
  left_join(dens_biomarkers, dens_loads) %>% 
  left_join(dens_univar_out) %>% 
  left_join(annotations) %>% 
  mutate_if(is.numeric, round, 3) %>% 
  #format correlations with astrixes instead of showing p-values
  mutate(correlation = case_when(
    corr.p.value <= 0.001 ~ paste0(correlation, "***"),
    corr.p.value <= 0.01  ~ paste0(correlation, "**"),
    corr.p.value <=0.05   ~ paste0(correlation, "*"),
    TRUE                  ~ as.character(correlation)
  )) %>% 
  select(Compound            = pretty_name,
         CAS                 = cas,
         "RDA Loading"       = RDA1,
         Correlation         = correlation,
         "Response type"     = type,
         "Response estimate" = estimate,
         "Change point"      = threshold,
         "Chemical Family"   = chemical_family,
          Aroma              = aroma)

write_excel_csv(table1, here("figs", "2017 density biomarkers.csv"), na = "-")
```

## Biomarkers for damage

### Extract scores and loadings

```{r}
dam_scores <- scores(rda_t, display = "sites", scaling = 0) %>% as_tibble(rownames = "sample")
dam_loads <- scores(rda_t, display = "species", scaling = 0) %>% as_tibble(rownames = "compound")
```

### Correlation with RDA axis

```{r}
dam_corr <-
  map_df(chemdata,
         ~cor.test(.x, dam_scores$RDA1) %>% 
           broom::glance(),
         .id = "compound") %>% 
  select(compound, correlation = estimate, corr.p.value = p.value)
```

### Filter

```{r}
dam_biomarkers <-
  dam_corr %>% 
  filter(corr.p.value <= 0.05) %>% 
  arrange(desc(abs(correlation)))
```
### Univariate regressions


```{r}
#Only need to do this on compounds in biomarkers list
dam_biom_data <-
  gc_tidy %>% 
  mutate(log.RPA = log(RPA)) %>% 
  filter(compound %in% dam_biomarkers$compound)
```


```{r}
dam_univar <-
  dam_biom_data %>% 
  group_by(compound) %>% 
  nest() %>% 
  #fit null, linear, step, and hing models
  mutate(null = map(data, ~lm(log.RPA ~ 1, data = .x)),
         line = map(data, ~lm(log.RPA ~ log.twister, data = .x)),
         step = map(data, ~chngptm(log.RPA ~ 1, ~ log.twister,
                                   family = "gaussian", type = "step", data = .x, var.type = "bootstrap", ci.bootstrap.size = 100)),
         hinge = map(data, ~chngptm(log.RPA ~ 1, ~ log.twister, 
                                    family = "gaussian", type = "hinge", data = .x))) %>% 
  gather(-compound, -data, key = type, value = model) %>% 
  #extract winning glm from step and hing models
  mutate(fit = ifelse(type %in% c("step", "hinge"), map(model, ~.x$best.fit), model)) %>% 
  arrange(compound) %>% 
  mutate(AIC = map(fit, ~AIC(.x))) %>%
  unnest(AIC)
```

Get model info and coeficients and choose the best model by AIC, penalizing any hinge or step models that have threshold CIs that overlap zero.

```{r}
# This is hacky, but works.  Ran into error when trying to do it all in one dataset.
dam_univar2 <- 
  dam_univar %>% 
  filter(type %in% c("step", "hinge")) %>% 
  mutate(coefs = map(model, tidy.chngpt)) %>% 
  unnest(coefs) %>% 
  select(-data, -model, -fit) %>% 
  #joins to null and line models
  left_join(dam_univar, .) %>%
  #set AIC to very high if the threshold CI overlaps zero
  mutate(AIC = if_else(chngpt.lower.CI > 0 | is.na(chngpt.lower.CI), AIC, Inf)) %>%
  mutate(AIC = if_else(chngpt.upper.CI < max(gc_tidy$log.twister) | is.na(chngpt.upper.CI),
                       AIC, Inf)) %>% 
  #set AIC to very high if coefficient isn't significantly different from zero
  mutate(AIC = if_else(coef.p.value > 0.05, Inf, AIC,missing = AIC)) %>% 
  group_by(compound) %>%
  # choose the best AIC as described above
  filter(AIC == bestAIC(AIC))

# dam_univar %>% filter(compound =="Isovaleric acid", type == "step") %>% pull(model)
```

```{r}
dam_univar_out <-
  dam_univar2 %>% 
  mutate(estimate = case_when(type == "step"  ~ `I(log.twister>chngpt)`,
                              type == "hinge" ~ `(log.twister-chngpt)+`,
                              type == "line"  ~ map_dbl(fit, ~coef(.x)[2]))) %>% 
  select(compound, type, estimate, threshold = chngpt)
```

### Merge to create Table 2

Biomarker table should have: 

- Compound (pretty name)
- CAS
- RDA loading
- Correlation (with *** for p-values)
- Response type (null, linear, hinge, step)
- Response estimate (linear = slope, hinge = slope after threshold, step = intercept2 - intercept1)
- Change point
- Chemical family (???)
- Aroma

```{r}
table2 <-
  left_join(dam_biomarkers, dam_loads) %>% 
  left_join(dam_univar_out) %>% 
  left_join(annotations) %>% 
  mutate_if(is.numeric, round, 3) %>% 
  #format correlations with astrixes instead of showing p-values
  mutate(correlation = case_when(
    corr.p.value <= 0.001 ~ paste0(correlation, "***"),
    corr.p.value <= 0.01  ~ paste0(correlation, "**"),
    corr.p.value <=0.05   ~ paste0(correlation, "*"),
    TRUE                  ~ as.character(correlation)
  )) %>% 
  select(Compound            = pretty_name,
         CAS                 = cas,
         "RDA Loading"       = RDA1,
         Correlation         = correlation,
         "Response type"     = type,
         "Response estimate" = estimate,
         "Change point"      = threshold,
         "Chemical Family"   = chemical_family,
          Aroma              = aroma)

write_excel_csv(table2, here("figs", "2017 damage biomarkers.csv"), na = "-")
```
# Supplemental table

Supplemental table should have:

- Compound
- Unknowns re-nammed by RI order
- CAS
- experimental RI
- Lit/lab RI
- Correlation with density RDA
- Correlation with damage RDA

## Join Loadings
```{r}
all_loads <-
  full_join(dens_loads, dam_loads, by = "compound") %>% 
  select(compound, density_rda_load = RDA1.x, damage_rda_load = RDA1.y)
```

## Join Correlations
```{r}
all_cors <-
  full_join(dens_corr, dam_corr, by = "compound") %>% 
  select(compound,
         density_rda_corr = correlation.x, density_corr_pval = corr.p.value.x,
         damage_rda_corr = correlation.y, damage_corr_pval = corr.p.value.y)
```

## join loadings and correlations

```{r}
all_rda_output <-
  full_join(all_loads, all_cors, by = "compound") %>% 
  select(compound, starts_with("density"), starts_with("damage"))
```

## Join with annotations and clean up
```{r}
s_table <-
  full_join(all_rda_output, annotations, by = "compound") %>% 
  mutate(RI_lab_lit = ifelse(is.na(ri_lab), ri_lit, ri_lab),
         source2 = ifelse(!is.na(ri_lab), "standard", source)) %>% 
  select(Compound = pretty_name, CAS = cas, RI = meanRI, RI_lab_lit, Source = source2,
         starts_with("density"), starts_with("damage")) %>% 
  arrange(RI)
```

## Save!

```{r}
write_csv(s_table, here("figs", "TableS1.csv"))
```

# Plots


## Density Biomarkers

```{r}
dens_plotdata <- 
  full_join(dens_biomarkers, dens_univar2, by = "compound") %>% 
  left_join(annotations, by = "compound") %>%
  dplyr::slice(1:9) %>% # Just the first 9
  left_join(dens_biom_data, by = "compound") %>% 
  mutate(pretty_name = fct_reorder(pretty_name, abs(correlation), mean, .desc = TRUE))

#just linear models
dens_justlm <- dens_plotdata %>% filter(type == "line")
```

```{r}
dens_plot <- 
  ggplot(dens_plotdata, aes(x = density_end, y = log.RPA)) +
  geom_point(alpha = 0.5) +
  geom_segment(aes(x = min(density_end),
                 xend = chngpt,
                 y = `(Intercept)`,
                 yend = `(Intercept)`), color = "black") +
  #step
  geom_segment(aes(x = chngpt,
                   xend = max(density_end),
                   y = `(Intercept)`+`I(density_end>chngpt)`, 
                   yend = `(Intercept)`+`I(density_end>chngpt)`), color = "black") +
  #hinge
  geom_segment(aes(x = chngpt,
                   xend = max(density_end),
                   y = `(Intercept)`,
                   yend = `(Intercept)` + `(density_end-chngpt)+`)) +
  #line
  geom_smooth(method = "lm", se = FALSE, color = "black", data = dens_justlm, size = 0.5) +
  facet_wrap(~pretty_name, scales = "free_y") +
  labs(x = "Leafhopper density (insects / leaf)",
       y = "ln(relative peak area)") +
  theme_bw()
dens_plot
```


```{r}
save_plot(here("figs", "density-biomarkers.png"), dens_plot,
          ncol = 3,
          nrow = 3, 
          base_height = 2,
          base_asp = 1.3)
```


## Damage biomarkers

```{r}
dam_plotdata <- 
  full_join(dam_biomarkers, dam_univar2, by = "compound") %>% 
  left_join(annotations, by = "compound") %>%
  dplyr::slice(1:9) %>% 
  left_join(dam_biom_data, by = "compound") %>% 
  mutate(pretty_name = fct_reorder(pretty_name, abs(correlation), mean, .desc = TRUE))

#just linear models
dam_justlm <- dam_plotdata %>% filter(type == "line")
```

```{r}
dam_plot <- 
  ggplot(dam_plotdata, aes(x = log.twister, y = log.RPA)) +
  geom_point(alpha = 0.5) +
  geom_segment(aes(x = min(log.twister),
                 xend = chngpt,
                 y = `(Intercept)`,
                 yend = `(Intercept)`), color = "black") +
  #step
  geom_segment(aes(x = chngpt,
                   xend = max(log.twister),
                   y = `(Intercept)`+`I(log.twister>chngpt)`, 
                   yend = `(Intercept)`+`I(log.twister>chngpt)`), color = "black") +
  #hinge
  geom_segment(aes(x = chngpt,
                   xend = max(log.twister),
                   y = `(Intercept)`,
                   yend = `(Intercept)` + `(log.twister-chngpt)+` * (max(gc_tidy$log.twister) - chngpt))) +
  #line
  geom_smooth(method = "lm", se = FALSE, color = "black", data = dam_justlm, size = 0.5) +
  facet_wrap(~pretty_name, scales = "free_y") +
  labs(x = "ln(% damage to focal leaf)",
       y = "ln(relative peak area)") +
  theme_bw()
dam_plot
```


```{r}
save_plot(here("figs", "damage-biomarkers.png"), dam_plot,
          ncol = 3,
          nrow = 3,
          base_height = 2,
          base_asp = 1.3)
```
