---
title: "2018 Leafhopper GCMS analysis"
author: "Eric R. Scott"
date: "2019-08-01"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Last compiled: `r Sys.Date()`*

```{r packages, include=FALSE}
library(tidyverse)
library(here)
library(readxl)
#for RDA:
library(vegan)
library(RVAideMemoire)
#for figures:
library(cowplot)
#for step and hinge models
library(chngpt)
```

# Purpose

This is the main analysis of volatile data for the paper. Multivariate analysis with redundancy analysis (RDA).  I decided to go with RDA over PLSR because I like that it gives me the percentage of variation explained by the predictor variables (herbivory proxies) whereas PLSR just gives a R2 for the axis (how much variation in X and Y does the *axis* explain).  I then identify biomarkers as compounds with significant correlations to the first (only) RDA axis, and do some univariate tests on them to see if they best fit a null (intercept only), line, step function, or hinge and then plot the results and produce a table.

# Load Data
Read in cleaned GCMS data (which includes leafhopper density already) and leafhopper damage data (from image analysis)
```{r data, echo=TRUE}
gc_wide <- read_rds(here("data", "cleaned", "2018_gcms_wide.rds"))
gc_tidy <- read_rds(here("data", "cleaned", "2018_gcms_tidy.rds")) %>% rename(compound = Compound)
# gc_wide.zeroes <- read_rds(here("data", "cleaned", "2018_gcms_zeroes.rds"))
```

```{r}
annotations.raw <- 
  read_excel(here("data", "cleaned", "Compound annotation updated.xlsx"), na = "NA") %>%
  janitor::clean_names()
```

## Data Dictionary

`gc_tidy`: A tidy dataframe of all the chemistry and other data

- `sample`: Sample name
- `cultivar`: Either J (Jin Guan Yin) or L (Long Jing #43)
- `density_start`: The density of leafhoppers applied at the start of the experiment in **insects/shoot**
- `density_end`: Ending density in **insects/shoot** 
- `mean_percent_damage`: Percentage of damaged pixels averaged across all leaves for that sample
- `twister_damage`: Percentage of damaged pixels for the leaf the volatiles were sampled from. Many values are missing because I forgot to mark which leaf had the Twister on it.
- `No.`: Compound number.  This comes from the Ion Analytics methods file.
- `Compound`: Compound name
- `RPA`: Relative peak area
- `rt`: retention time, in minutes
- `ri`: retention index
- `present`: logical. Was the compound detected in a particular sample?

`gc_wide`: A wide version with columns for each `Compound` with `RPA` as the value

# Load Functions
Custom functions needed for analysis:

```{r}
source(here("R", "functions.R"))
```

# Data pre-treatment
I'm going to log-transform and scale the wide data

```{r}
metavars <- c("sample", "cultivar", "density_start", "density_end", "mean_percent_damage", "twister_damage")
```
 
```{r}
gc_wide.logscale <- 
  gc_wide %>% 
  mutate_at(vars(-metavars), log) %>%
  mutate_at(vars(-metavars), scale) %>% 
  #also log-transform leaf damage
  mutate(log.twister = log(twister_damage),
         log.mean_damage = log(mean_percent_damage)) %>% 
  #strip atrributes left by scale() that interfere with some other functions down the road
  mutate_at(vars(-metavars), rm_attr)
```


# Analysis
# Descriptive statistics
1. How many compounds?
```{r}
length(unique(gc_tidy$compound))
```
2. How many in Long Jing, how many in JGY?
```{r}
gc_tidy %>%
  filter(cultivar == "J") %>% 
  group_by(compound) %>% #for each compound...
  filter(sum(present) > 0) %>% group_by(sample) %>% tally()

gc_tidy %>% 
  filter(cultivar == "L") %>% 
  group_by(compound) %>% 
  filter(sum(present) > 0) %>%
  group_by(sample) %>% tally()
```

3. How many found in **all** JGY or **all** LJ samples?
```{r}
jgy_common <- 
  gc_tidy %>% 
  filter(cultivar == "J") %>% #count() #22 samples total
  summarize(nsamp = sum(present)) %>% 
  filter(nsamp == 20)
nrow(jgy_common)
# jgy_common
```

```{r}
lj_common <- 
  gc_tidy %>% 
  filter(cultivar == "L") %>% #count() #19 samples total
  summarize(nsamp = sum(present)) %>% 
  filter(nsamp == 19)
nrow(lj_common)
# lj_common
```

# Analysis
Because the effect of cultivar is confounded by the time of the experiment, and because preliminary analysis showed there were highly significant differences between cultivars, analysis procedes with cultivars separated.  Additionally, the peak areas of the internal standard are an order of magnitude different between cultivars.

I need to split them *before* scaling!

```{r}
gc_jgy <-   
  gc_wide %>% 
  filter(cultivar == "J") %>% 
  mutate_at(vars(-metavars), log) %>%
  mutate_at(vars(-metavars), scale) %>% 
  #strip atrributes left by scale() that interfere with some other functions down the road
  mutate_at(vars(-metavars), rm_attr) %>% 
  mutate(log.twister = log(twister_damage),
         log.mean_damage = log(mean_percent_damage))
gc_jgy.t <-   
  gc_wide %>% 
  filter(cultivar == "J") %>% 
  filter(!is.na(twister_damage)) %>% 
  mutate_at(vars(-metavars), log) %>%
  mutate_at(vars(-metavars), scale) %>% 
  #strip atrributes left by scale() that interfere with some other functions down the road
  mutate_at(vars(-metavars), rm_attr) %>% 
  mutate(log.twister = log(twister_damage),
         log.mean_damage = log(mean_percent_damage))
gc_lj <- 
  gc_wide %>% 
  filter(cultivar == "L") %>% 
  mutate_at(vars(-metavars), log) %>%
  mutate_at(vars(-metavars), scale) %>% 
  #strip atrributes left by scale() that interfere with some other functions down the road
  mutate_at(vars(-metavars), rm_attr) %>% 
  mutate(log.twister = log(twister_damage),
         log.mean_damage = log(mean_percent_damage))
gc_lj.t <- 
  gc_wide %>% 
  filter(cultivar == "L") %>% 
  filter(!is.na(twister_damage)) %>%
  mutate_at(vars(-metavars), log) %>%
  mutate_at(vars(-metavars), scale) %>% 
  #strip atrributes left by scale() that interfere with some other functions down the road
  mutate_at(vars(-metavars), rm_attr) %>% 
  mutate(log.twister = log(twister_damage),
         log.mean_damage = log(mean_percent_damage))

#update list of variables that aren't RPAs
metavars <- c(metavars, "log.twister", "log.mean_damage")
```

## Jin Guan Yin
### Fit RDAs
fit RDAs using leafhopper density, mean leaf damage, and twister leaf damage (using subset of samples with twister leaf data)
```{r}
rda_jgy_dens <- 
  rda(gc_jgy %>% select(-metavars) ~ density_end,
      data = gc_jgy)

rda_jgy_dam <-
  rda(gc_jgy %>% select(-metavars) ~ log.mean_damage,
      data = gc_jgy)

rda_jgy_t <- 
  rda(gc_jgy.t %>% select(-metavars) ~ log.twister,
      data = gc_jgy.t)
```

### Variance Explained
```{r}
MVA.synt(rda_jgy_dens)[[1]][[2]]
```
```{r}
MVA.synt(rda_jgy_dam)[[1]][[2]]
```
```{r}
MVA.synt(rda_jgy_t)[[1]][[2]]
```

All proxies of herbivory explain < 6% of variation in volatiles

### Anovas

```{r}
anova(rda_jgy_dens)
```

```{r}
anova(rda_jgy_dam)
```

```{r}
anova(rda_jgy_t)
```

All non-significant

## Long Jing
### Fit RDAs for Long Jing

```{r}
rda_lj_dens <-
  rda(gc_lj %>% select(-metavars) ~ density_end,
      data = gc_lj)

rda_lj_dam <-
  rda(gc_lj %>% select(-metavars) ~ log.mean_damage,
      data = gc_lj)

rda_lj_t <-
  rda(gc_lj.t %>% select(-metavars) ~ log.twister,
      data = gc_lj.t)
```

### Variance explained

```{r}
MVA.synt(rda_lj_dens)[[1]][[2]]
```

```{r}
MVA.synt(rda_lj_dam)[[1]][[2]]
```

```{r}
MVA.synt(rda_lj_t)[[1]][[2]]
```

Twister leaf damage explains the most variation, but this is also a very small subset of the data (only 7 samples)

### ANOVAs
```{r}
anova(rda_lj_dens)
```

```{r}
anova(rda_lj_dam)
```

```{r}
anova(rda_lj_t)
```

None are significant

# Supplemental table
Renumber unknowns by RI order
```{r}
meanRIs <-
  gc_tidy %>% 
  group_by(compound) %>% 
  summarize(meanRI = mean(ri, na.rm = TRUE))

annotations2 <- left_join(meanRIs, annotations.raw, by = "compound")
  
unks <-
  annotations2 %>% 
  filter(is.na(cas)) %>% 
  arrange(meanRI) %>% 
  add_column(unk_name = 1:nrow(.))

annotations <- 
  left_join(annotations2, unks) %>% 
  mutate(pretty_name = ifelse(is.na(unk_name), pretty_name, unk_name)) %>% 
  mutate(pretty_name = ifelse(is.na(pretty_name), compound, pretty_name)) %>% 
  select(compound, cas, pretty_name, meanRI, ri_lab, ri_lit, source, aroma, chemical_family) %>% 
  mutate(chemical_family = snakecase::to_sentence_case(chemical_family))
```

Cleanup
```{r}
s_table<-
  annotations %>% 
  mutate(RI_lab_lit = ifelse(is.na(ri_lab), ri_lit, ri_lab),
         source2 = ifelse(!is.na(ri_lab), "standard", source)) %>% 
  select(Compound = pretty_name, CAS = cas, RI = meanRI, RI_lab_lit, Source = source2) %>% 
  arrange(RI) %>% 
  mutate_if(is.double, round, 0)
s_table
```


Write it

```{r}
write_csv(s_table, here("figs", "TableS2.csv"))
```

