---
title: "LCMS analysis"
author: "Eric R. Scott"
date: "2019-07-26"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# options(scipen =999) #turn off scientific notation
```

*Last compiled: `r Sys.Date()`*

```{r packages, include=FALSE}
library(tidyverse)
library(lubridate)
library(here)
library(vegan)
library(RVAideMemoire)
library(chemhelper)
library(naniar)
library(visdat)
library(broom)
library(modelr)
library(glue)
library(cowplot)
library(chngpt)

here <- here::here #make sure to override the here() function from lubridate
```
```{r}
source(here("R", "functions.R"))
```

# Purpose

Analysis of LCMS data from 2017 and 2018 leafhopper density experiments.

# Load Data

```{r data, echo=TRUE}
lc2017 <- read_rds(here("data", "cleaned", "2017_lcms_wide.rds")) %>% 
  filter(sample != "Q2") #determined elsewhere that this has an unusually high ammount of damage for the treatment

lc2018 <- read_rds(here("data", "cleaned", "2018_lcms_wide.rds")) %>% 
  filter(sample != "J23") #determined elswhere that this has an unusually high ammount of damage for the treatment

lc2017_tidy <-
  read_rds(here('data', 'cleaned', '2017_lcms_tidy.rds')) %>% 
  filter(sample != "Q2")

lc2018_tidy <- 
  read_rds(here('data', 'cleaned', '2018_lcms_tidy.rds')) %>% 
  filter(sample != "J23")
```

## Data Dictionary

`lc2017` and `lc2018`: Wide versions of LCMS data for both years with compounds as columns.

- `sample`: sample name
- `cultivar`: Q = Qing Xin Da Mao, L = Long Jing, J = Jin Guan Yin
- `plant_num`: another identifier, extracted from sample name
- `mean_leaf_area_px`: mean leaf area in pixels
- `mean_percent_damage`: mean percent area damaged
- `n_leaves`: number of leaves scanned and ground into powder for this analysis
- `twister_damage`: percent damage on the twister leaf.  Ignore for LCMS analysis.
- `density_treatmetn`: categorical.  This was the leafhopper density I aimed to treat the plants with in insects/young leaf for 2017 and insects/shoot in 2018
- `density_start`, `density_end`: The density of leafhoppers at the start and end of the experiments.  In insects/young leaf in 2017 and insects/shoot in 2018
- The rest of the columns are concentrations of compounds in **µg/mg** leaf

# Analysis
## Deal with NAs

NAs are non-detects.  I should replace them with 0 or a very small number.

```{r}
vis_dat(lc2017)
vis_dat(lc2018)
```


```{r}
# min(lc2017_tidy$conc_ug_mg, na.rm = TRUE)

lc2017_tidy <- 
  lc2017_tidy %>% 
  filter(!compound %in% c("Catechin gallate", "Theophylline")) %>% #not detected in any sample
  mutate(conc_ug_mg = ifelse(is.na(conc_ug_mg), 0.0001, conc_ug_mg))

lc2017_2 <-
  lc2017 %>% 
  mutate_all(~ifelse(is.na(.), 0.0001, .)) %>% 
  select(-`Catechin gallate`, -`Theophylline`) #all zeroes

# min(lc2018_tidy$conc_ug_mg, na.rm = TRUE)
lc2018_tidy <-
  lc2018_tidy %>% 
  filter(!compound %in% c("Catechin gallate", "Theophylline")) %>% #all zeros for all samples
  mutate(conc_ug_mg = ifelse(is.na(conc_ug_mg), 0.0001, conc_ug_mg)) 

lc2018_2 <- 
  lc2018 %>% 
  mutate_all(~ifelse(is.na(.), 0.0001, .)) %>% 
  select(-`Catechin gallate`, -Theophylline) #all zeroes
```

## Scale and transform data

```{r}
lc2017_scaled <-
  lc2017_2 %>%
  mutate_at(vars(Caffeine:Theobromine), scale) %>% 
  mutate(log_damage = log(mean_percent_damage))

lc2017_tidy <- 
  lc2017_tidy %>% 
  mutate(log_damage = log(mean_percent_damage))

lc2018_2 <- 
  lc2018_2 %>% 
  mutate(log_damage = log(mean_percent_damage))

lc2018_tidy <- 
  lc2018_tidy %>% 
  mutate(log_damage = log(mean_percent_damage))
```

## Examine correlations

```{r}
corrr::correlate(lc2017_2[11:21]) %>% 
  corrr::shave() %>% 
  corrr::rplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "2017")

corrr::correlate(lc2018_2[11:21]) %>% 
  corrr::shave() %>% 
  corrr::rplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "2018")
```

There are some strong correlations, so multivariable methods are best.

# 2017
## Leafhopper density RDA

1. Fit RDA model

```{r}
dens_rda.2017 <-
  rda(lc2017_scaled %>% select(Caffeine:Theobromine) ~ density_end, data = lc2017_scaled)
```


2. How much total variance does the experimental design explain?

```{r}
MVA.synt(dens_rda.2017)[[1]][[2]]
```

Only 5.45% variation explained

3. Test for significance that this amount of explained variance is higher than the null hypothesis of no effect of the experimental design:

```{r}
anova(dens_rda.2017)
```

## Mean leaf damage RDA

1. fit model
```{r}
dam_rda.2017 <-
  rda(lc2017_scaled %>% select(Caffeine:Theobromine) ~ log_damage, data = lc2017_scaled)
```


2. How much total variance does the experimental design explain?

```{r}
MVA.synt(dam_rda.2017)[[1]][[2]]
```

9.58 % of variation explained

3. Test for significance that this amount of explained variance is higher than the null hypothesis of no effect of the experimental design:

```{r}
anova(dam_rda.2017)
```
Marginally significant effect of leaf damage on metabolites.

### Get biomarkers for damage

```{r}
scores <- MVA.scores(dam_rda.2017)$coord
loads <- MVA.load(dam_rda.2017)$loads %>% rename(loading = `Constr. comp. 1`)
data <- lc2017_scaled %>% select(Caffeine:Theobromine)

lc_biomarkers_2017 <-
  map_df(data, ~cor.test(.x, scores[["Constr. comp. 1"]]) %>% broom::glance(), .id = "compound") %>%
  bind_cols(loads) %>%
  filter(p.value <= 0.05) %>%
  select(compound, loading, correlation = estimate, p.value) %>% 
  arrange(desc(abs(correlation)))
lc_biomarkers_2017
```


Top contributors: EGCG, Caffeine, Theobromine, ECG.  All negative relationship with mean % damage.

Univariate measures

Fit null, linear, step, and hinge models to all four compounds.

```{r}
lc_lm_2017 <-
  left_join(lc_biomarkers_2017, lc2017_tidy) %>% 
  mutate(log_damage = log(mean_percent_damage)) %>% 
  group_by(compound) %>% 
  nest() %>% 
  mutate(linear = map(data, ~lm(conc_ug_mg ~ log_damage, data = .x)),
         null = map(data, ~lm(conc_ug_mg ~ 1, data = .x)),
         step = map(data, ~ chngptm(conc_ug_mg ~ 1, ~ log_damage,
                                    family = "gaussian", type = "step", data = .x, 
                                    var.type = "bootstrap", ci.bootstrap.size = 100)),
         hinge = map(data, ~chngptm(conc_ug_mg ~ 1, ~ log_damage,
                                    family = "gaussian", type = "hinge", data = .x))) %>%
  gather(-compound, -data, key = type, value = model) %>% 
  mutate(fit = ifelse(type %in% c("step", "hinge"), map(model, ~.x$best.fit), model)) %>% 
  arrange(compound) %>% 
  mutate(AIC = map(fit, ~AIC(.x))) %>%
  unnest(AIC)


lc_lm_2017 <-
  lc_lm_2017 %>% 
  filter(type %in% c("step", "hinge")) %>% 
  mutate(coefs = map(model, tidy.chngpt)) %>% 
  unnest(coefs) %>% 
  select(-data, -model, -fit) %>% 
  #joins to null and line models
  left_join(lc_lm_2017, .) %>%
  #set AIC to very high if the threshold CI overlaps zero
  mutate(AIC = if_else(chngpt.lower.CI > min(lc2017_scaled$log_damage) | is.na(chngpt.lower.CI), AIC, Inf)) %>%
  mutate(AIC = if_else(chngpt.upper.CI < max(lc2017_scaled$log_damage) | is.na(chngpt.upper.CI), AIC, Inf)) %>% 
  group_by(compound) 

lc_lm_2017
# lc_lm_2017$model[[13]] %>% summary()

```


Choose best fit manually by AIC:

- Caffeine: null
- Epicatechin gallate: step is best fit, but not significant, so null
- Epigallocatechin gallate: linear
- Theobromine: step is best fit, but not significant, same with hinge, linear is marginally significant

Check that changepoint doesn't overlap zero

```{r}
lc_lm_2017 <-
  lc_lm_2017 %>% 
  ungroup() %>% 
  filter(compound == "Caffeine" & type == "null" |
           compound == "Epicatechin gallate" & type == "null" |
           compound == "Epigallocatechin gallate" & type == "linear" |
           compound == "Theobromine" & type == "linear")
```



```{r}
lc_biomarkers_2017 <- full_join(lc_biomarkers_2017, lc_lm_2017)
```

## Plots

```{r}
lc_plotdata.2017 <-
  left_join(lc_biomarkers_2017, lc2017_tidy) %>% 
  mutate(compound = fct_reorder(compound, abs(loading), .desc = TRUE))
  
lc_plot.2017 <-
  ggplot(lc_plotdata.2017, aes(x = log_damage, y = conc_ug_mg)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm",
              se = FALSE,
              data = lc_plotdata.2017 %>% 
                filter(compound %in% c("Epigallocatechin gallate", "Theobromine")),
              color = "black", size = 0.5, linetype = 1) +
  #  #step
  # geom_segment(aes(x = min(log_damage),
  #                xend = chngpt,
  #                y = `(Intercept)`,
  #                yend = `(Intercept)`), color = "black") +
  # geom_segment(aes(x = chngpt,
  #                  xend = max(log_damage),
  #                  y = `(Intercept)`+`I(log_damage>chngpt)`, 
  #                  yend = `(Intercept)`+`I(log_damage>chngpt)`), color = "black") +
  facet_wrap(~compound, scales = "free_y") +
  theme_bw() +
  labs(x = "ln(mean % leaf damage)",
       y = "Concentration (µg/mg)")
lc_plot.2017

lc_plotdata.2017 %>% count(sample)
```

```{r}
save_plot(here("figs", "lc_biomarker_plot_2017.png"), lc_plot.2017,
          ncol = 2,
          nrow = 2,
          base_height = 2,
          base_asp = 1.3)
save_plot(here("figs", "figure5.jpg"), lc_plot.2017,
          ncol = 2,
          nrow = 2,
          base_height = 2,
          base_asp = 1.3)
```

# 2018


## Checking effect of processing time
In this year, I was unable to keep samples refridgerated as I processed them (scanning, then microwaving, then drying).  Scanning was the bottleneck step, so the first samples spent a lot less time sitting out on the counter after being harvested than the last samples.  To check for an effect of sample order, I pulled the file creation times from all the scanned images and merged them with the data.  Then, I used the time the scan was created as a conditional variable in an RDA.  It only explained about 1% of the variation in the LCMS compounds, so I won't use it as a covariate going forward.

```{r eval=FALSE}
jgy_path <- "/Volumes/as_rsch_orianslab_tea01$/Image Analysis/2018/Leaf Scans/Manipulative/Jin Guan Yin"
lj_path <- "/Volumes/as_rsch_orianslab_tea01$/Image Analysis/2018/Leaf Scans/Manipulative/Longjing"

jgy_info <-
  dir(jgy_path, "*.jpg", full.names = TRUE) %>%
  file.info() %>% 
  as_tibble(rownames = "paths") %>% 
  mutate(plant_num = str_extract(paths, "\\d+(?=\\w?\\.jpg$)")) %>% 
  mutate(sample = paste0("J", plant_num)) %>% 
  mutate(mtime = if_else(sample == "J18",
                         ymd_hms("2018-07-30 23:17:00") + hours(4),
                         mtime)) %>% 
  mutate(time = mtime - min(mtime))

lj_info <-
  dir(lj_path, "*.png", full.names = TRUE) %>% 
  file.info() %>% 
  as_tibble(rownames = "paths") %>% 
  mutate(plant_num = str_extract(paths, "\\d+(?=\\w?\\.png$)")) %>% 
  mutate(sample = paste0("L", plant_num)) %>% 
  mutate(time = mtime - min(mtime))

processing_time <-
  bind_rows(jgy_info, lj_info) %>% 
  select(sample, plant_num, time) %>% 
  group_by(sample) %>% 
  summarize(time = min(time))
```

```{r eval=FALSE}
lc2018_scaled <- 
  lc2018_2 %>% 
  mutate_at(vars(Caffeine:Theobromine), scale) %>% 
  mutate(cultivar = as.factor(cultivar)) %>% 
  left_join(processing_time)

rda.2018 <- 
  rda(lc2018_scaled %>% select(Caffeine:Theobromine) ~ log_damage*cultivar + Condition(as.numeric(time)),
      data = lc2018_scaled)
```

```{r eval=FALSE}
MVA.synt(rda.2018)
```

## Split cultivars

It doesn't make sense to include both cultivars in the analysis.

```{r}
lc_jgy <- 
  lc2018_2 %>% 
  filter(cultivar == "J") %>% 
  mutate_at(vars(Caffeine:Theobromine), scale)
lc_lj <-
  lc2018_2 %>% 
  filter(cultivar == "L") %>% 
  mutate_at(vars(Caffeine:Theobromine), scale)
```


## Jin Guan Yin

1. Fit RDA models

```{r}
rda_dens_jgy <- 
  rda(lc_jgy %>% select(Caffeine:Theobromine) ~ density_end,
      data = lc_jgy)

rda_dam_jgy <- 
  rda(lc_jgy %>% select(Caffeine:Theobromine) ~ log_damage,
      data = lc_jgy)
```

2. How much total variance does the experimental design explain?

```{r}
MVA.synt(rda_dens_jgy)[[1]][[2]]
MVA.synt(rda_dam_jgy)[[1]][[2]]
```
Very small numbers here.

3. Test for significance that this amount of explained variance is higher than the null hypothesis of no effect of the experimental design:

```{r}
anova(rda_dens_jgy)
anova(rda_dam_jgy)
```
Definitely not significant

## Long Jing

1. Fit RDA model

```{r}
rda_dens_lj <- 
  rda(lc_lj %>% select(Caffeine:Theobromine) ~ density_end,
      data = lc_lj)

rda_dam_lj <- 
  rda(lc_lj %>% select(Caffeine:Theobromine) ~ log_damage,
      data = lc_lj)
```

2. How much total variance does the experimental design explain?

```{r}
MVA.synt(rda_dens_lj)[[1]][[2]]
MVA.synt(rda_dam_lj)[[1]][[2]]
```
~ 6-7% for both

3. Test for significance that this amount of explained variance is higher than the null hypothesis of no effect of the experimental design:

```{r}
anova(rda_dens_lj)
anova(rda_dam_lj)
```

not significant