---
title: "LCMS analysis"
author: "Eric R. Scott"
date: "2019-07-26"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Last compiled: `r Sys.Date()`*

```{r packages, include=FALSE}
library(tidyverse)
library(here)
library(ropls)
library(chemhelper)
library(naniar)
library(visdat)
```

# Purpose

Analysis of LCMS data from 2017 and 2018 leafhopper density experiments.

# Load Data

```{r data, echo=TRUE}
lc2017 <- read_rds(here("data", "cleaned", "2017_lcms_wide.rds")) %>% 
  filter(sample != "Q2") #determined elsewhere that this has an unusually high ammount of damage for the treatment
lc2018 <- read_rds(here("data", "cleaned", "2018_lcms_wide.rds"))
```

## Data Dictionary

`lc2017` and `lc2018`: Wide versions of LCMS data for both years with compounds as columns.

- `sample`: sample name
- `cultivar`: Q = Qing Xin Da Mao, L = Long Jing, J = Jin Guan Yin
- `plant_num`: another identifier, extracted from sample name
- `mean_leaf_area_px`: mean leaf area in pixels
- `mean_percent_damage`: mean percent area damaged
- `n_leaves`: number of leaves scanned and ground into powder for this analysis
- `twister_damage`: percent damage on the twister leaf.  Ignore for LCMS analysis.
- `density_treatmetn`: categorical.  This was the leafhopper density I aimed to treat the plants with in insects/young leaf for 2017 and insects/shoot in 2018
- `density_start`, `density_end`: The density of leafhoppers at the start and end of the experiments.  In insects/young leaf in 2017 and insects/shoot in 2018
- The rest of the columns are concentrations of compounds in **µg/mg** leaf

# Analysis
## Deal with NAs

NAs are non-detects.  I should replace them with 0 or a very small number.

```{r}
vis_dat(lc2017)
```


```{r}
lc2017_2 <- lc2017 %>% 
  mutate_all(~ifelse(is.na(.), 0.0001, .)) %>% 
  select(-`Catechin gallate`) #all zeroes
```

## Examine correlations

```{r}
corrr::correlate(lc2017_2[11:23]) %>% 
  corrr::shave() %>% 
  corrr::rplot() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

There are some strong correlations, so multivariate methods are probably best.

## 2017 PCA

```{r}
X <- lc2017_2[11:23]

opls(X, printL = FALSE)
```

## 2017 PLSR

No significant relationship between density or damage and metabolites. High R2Y, but very low Q2.  Overfitting?  Maybe if I remove some of the metabolites with a LOT of NAs?

```{r}
try(opls(X, lc2017_2$density_end))
opls(X, lc2017_2$mean_percent_damage, permI = 500, plotL = FALSE, printL = FALSE)
```



## Regression analysis

Ok, new approach.  Let's just do linear regressions?  Best way would be to load in the **tidy** versions of the data, do the same filtering (remove Q2, etc.), and then use `purrr` or something from `tidymodels` to do all the regressions.

```{r}
lc2017_2

caf <- lm(Caffeine ~ mean_percent_damage, data = lc2017_2)
summary(caf)
car::Anova(caf)

c <- lm(Catechin ~ mean_percent_damage, data = lc2017_2)
car::Anova(c)


```

