---
title: "LCMS analysis"
author: "Eric R. Scott"
date: "2019-07-26"
output: 
  html_notebook: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen =999) #turn of scientific notation
```

*Last compiled: `r Sys.Date()`*

```{r packages, include=FALSE}
library(tidyverse)
library(here)
library(vegan)
library(RVAideMemoire)
library(chemhelper)
library(naniar)
library(visdat)
library(broom)
library(modelr)
library(glue)
library(cowplot)
```

# Purpose

Analysis of LCMS data from 2017 and 2018 leafhopper density experiments.

# Load Data

```{r data, echo=TRUE}
lc2017 <- read_rds(here("data", "cleaned", "2017_lcms_wide.rds")) %>% 
  filter(sample != "Q2") #determined elsewhere that this has an unusually high ammount of damage for the treatment
lc2018 <- read_rds(here("data", "cleaned", "2018_lcms_wide.rds")) %>% 
  filter(sample != "J23") #determined elswhere that this has an unusually high ammount of damage for the treatment


lc2017_tidy <-
  read_rds(here('data', 'cleaned', '2017_lcms_tidy.rds')) %>% 
  filter(sample != "Q2")

lc2018_tidy <- 
  read_rds(here('data', 'cleaned', '2018_lcms_tidy.rds')) %>% 
  filter(sample != "J23")

```

## Data Dictionary

`lc2017` and `lc2018`: Wide versions of LCMS data for both years with compounds as columns.

- `sample`: sample name
- `cultivar`: Q = Qing Xin Da Mao, L = Long Jing, J = Jin Guan Yin
- `plant_num`: another identifier, extracted from sample name
- `mean_leaf_area_px`: mean leaf area in pixels
- `mean_percent_damage`: mean percent area damaged
- `n_leaves`: number of leaves scanned and ground into powder for this analysis
- `twister_damage`: percent damage on the twister leaf.  Ignore for LCMS analysis.
- `density_treatmetn`: categorical.  This was the leafhopper density I aimed to treat the plants with in insects/young leaf for 2017 and insects/shoot in 2018
- `density_start`, `density_end`: The density of leafhoppers at the start and end of the experiments.  In insects/young leaf in 2017 and insects/shoot in 2018
- The rest of the columns are concentrations of compounds in **µg/mg** leaf

# Analysis
## Deal with NAs

NAs are non-detects.  I should replace them with 0 or a very small number.

```{r}
vis_dat(lc2017)
vis_dat(lc2018)
```


```{r}
min(lc2017_tidy$conc_ug_mg, na.rm = TRUE)

lc2017_tidy <- 
  lc2017_tidy %>% 
  filter(compound != "Catechin gallate") %>% #not detected in any sample
  mutate(conc_ug_mg = ifelse(is.na(conc_ug_mg), 0.0001, conc_ug_mg))

lc2017_2 <-
  lc2017 %>% 
  mutate_all(~ifelse(is.na(.), 0.0001, .)) %>% 
  select(-`Catechin gallate`) #all zeroes

min(lc2018_tidy$conc_ug_mg, na.rm = TRUE)
lc2018_tidy <-
  lc2018_tidy %>% 
  filter(!compound %in% c("Catechin gallate", "Theophylline")) %>% #all zeros for all samples
  mutate(conc_ug_mg = ifelse(is.na(conc_ug_mg), 0.0001, conc_ug_mg)) 

lc2018_2 <- 
  lc2018 %>% 
  mutate_all(~ifelse(is.na(.), 0.0001, .)) %>% 
  select(-`Catechin gallate`, -Theophylline) #all zeroes
```

## Examine correlations

```{r}
corrr::correlate(lc2017_2[11:23]) %>% 
  corrr::shave() %>% 
  corrr::rplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "2017")

corrr::correlate(lc2018_2[11:22]) %>% 
  corrr::shave() %>% 
  corrr::rplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "2018")
```

There are some strong correlations, so multivariable methods might be best?

# 2017
## RDA

Probably RDA is better since there are some correlations?


1. Fit RDA model

```{r}
lc2017_scaled <- lc2017_2 %>% mutate_at(vars(Caffeine:Theophylline), scale)
rda.2017 <- rda(lc2017_scaled %>% select(Caffeine:Theophylline) ~ mean_percent_damage, data = lc2017_scaled)
```


2. How much total variance does the experimental design explain?

```{r}
MVA.synt(rda.2017)
```

Only 11.62%

3. Test for significance that this amount of explained variance is higher than the null hypothesis of no effect of the experimental design:

```{r}
anova(rda.2017)
```

Marginally significant effect of leaf damage on metabolites

4. Then we test for significance of individual factors and interaction terms:

```{r}
MVA.anova(rda.2017)
```

Marginally significant effect of leaf damage on metabolites.

5. Loadings

```{r}
# MVA.get.loads(rda.2017)
rda.2017$CCA$v %>% 
  as_tibble(rownames = "compound") %>% 
  arrange(RDA1)
```

Top contributors: EGCG, Caffeine, Theobromine.  All negative relationship with mean % damage.

## Regression analysis

Best way would be to load in the **tidy** versions of the data, do the same filtering (remove Q2, etc.), and then use tidyverse magic do all the regressions.


```{r}
results <-
  lc2017_tidy %>% 
  select(sample, compound, conc_ug_mg, mean_percent_damage, density_end) %>% 
  group_by(compound) %>% 
  nest() %>% #creates list column of data for each compound
  mutate(lm = map(data, ~lm(conc_ug_mg ~ mean_percent_damage, .))) %>%  # apply a linear model to each compound
  mutate(poly = map(data, ~lm(conc_ug_mg ~ poly(mean_percent_damage, 2), .))) %>%  # apply a polynomial model
  mutate(null = map(data, ~lm(conc_ug_mg ~ 1, .))) %>% # apply a null model
  select(compound, data, lm, poly, null) %>% 
  gather(-compound, -data, key = model.type, value = model) %>% #gather up all the models into one column
  mutate(stats = map(model, glance)) %>% #extract model stats
  unnest(stats) %>%
  group_by(compound) %>% 
  arrange(compound, AIC) %>% #figure out best model
  mutate(dAIC =  nth(AIC, 3)- AIC,
         dDF = nth(df, 3) - df) %>% #calculate dAIC
  select(compound, model.type, dAIC, dDF, p.value, everything())

results
```

Looks like EGCG has a pretty strong significant negative relationship with damage.  Makes sense as EGCG would be converted into theaflavins and other polymerized compounds.

- Caffeine: linear, **p = 0.05**
- Catechin: polynomial, p = 0.09
- Epicatechin: polynomial, p = 0.16
- Epicatechin gallate: linear, p = 0.096
- Epigallocatechin: null
- Epigallocatechin gallate: linear, **p = 0.0003**
- Gallic acid: null
- Gallocatechin: null
- Gallocatechin gallate: linear, p = 0.086
- L-theanine: null
- Paraxanthine: null
- Theobromine: linear, **p = 0.05**
- Theophylline: null

### Plotting results

First, creating a dataframe with only the "winning" models:

```{r}
results2 <-
  results %>% 
  filter(
    compound == "Caffeine" & model.type == "lm" |
      compound == "Catechin" & model.type == "poly" |
      compound == "Epicatechin" & model.type == "poly" |
      compound == "Epicatechin gallate" & model.type == "lm" |
      compound == "Epigallocatechin" & model.type == "null" |
      compound == "Epigallocatechin gallate" & model.type == "lm" |
      compound == "Gallic acid" & model.type == "null" |
      compound == "Gallocatechin" & model.type == "null" |
      compound == "Gallocatechin gallate" & model.type == "lm" |
      compound == "L-theanine" & model.type == "null" |
      compound == "Paraxanthine" & model.type == "null" |
      compound == "Theobromine" & model.type == "lm" |
      compound == "Theophylline" & model.type == "null"
  )
```

#### plot with facet_wrap()

Using `facet_wrap()` and two `geom_smooth()`s.  Pros: easier to read code.  Cons: more difficult to annotate with p-values? Maybe a custom labeller function?

```{r}
df <- full_join(lc2017_tidy, select(results2, compound, model.type, p.value))
df <- df %>% 
  mutate(compound = case_when(compound == "Caffeine" ~ "Caffeine(*)",
                              compound == "Epigallocatechin gallate" ~ "Epigallocatechin gallate(***)",
                              compound == "Theobromine" ~ "Theobromine(*)",
                              TRUE ~ as.character(compound)))
lc2017_plot <-
  ggplot(df, aes(x = mean_percent_damage, y = conc_ug_mg)) +
  geom_point() +
  geom_smooth(method = "lm", data = filter(df, model.type == "lm")) +
  geom_smooth(method = "lm", formula = "y ~ poly(x,2)", data = filter(df, model.type == "poly")) +
  facet_wrap(~compound, scales = "free_y") +
  labs(x = "Mean % damaged leaf area",
       y = "Concentration (µg/mg)",
       subtitle = "2017",
       caption = " (*): p = 0.05\n(***): p < 0.001") +
  theme_bw() +
  theme(legend.position = "none")
lc2017_plot
save_plot(here('figs', 'lc2017_plot.png'), lc2017_plot, ncol = 4, nrow = 4, base_height = 2)
```

#### plot using plot_grid()

Making a list of plots and then using `plot_grid()`.  Pros: easier to annotate with p-values.  Cons: don't know how to do axis labels better, don't know how to remove lines from certain plots.

```{r}
#Augment results with predicted values for drawing curves/lines
results2 <-
  results2 %>%
  mutate(pred = map2(model, data, ~{
    ndf <- tibble(mean_percent_damage = seq(min(.y$mean_percent_damage),
                                            max(.y$mean_percent_damage),
                                            length.out = 50))
    augment(.x, newdata = ndf)
    }))

# map a plotting function to every compound
plotlist <-
  pmap(list(.x = results2$data, .y = results2$pred, .name = results2$compound, .p = results2$p.value), 
     function(.x, .y, .name, .p) {
       ggplot(.x, aes(x = mean_percent_damage, y = conc_ug_mg)) +
         geom_point() +
         geom_line(aes(x = mean_percent_damage, y = .fitted), data = .y) +
         labs(
           x = "Mean % damaged leaf area",
           y = "Concentration (µg/mg)",
           subtitle = .name,
           caption = glue("p = {round(.p, 4)}")) +
         theme_bw() #+
         # theme(axis.title = element_blank())
     })

# plot_grid(plotlist = plotlist) %>%
  # save_plot(here('figs', 'lc2017_plot2.png'), ., ncol = 4, nrow = 4, base_height = 2)
```

# 2018

## RDA
Probably RDA is better since there are some correlations?

1. Fit RDA model

Added plant_num (i.e. sample processing order) as a conditional variable.  This pulls the variation from sample order out before doing the RDA---kind of like a random variable, but still a fixed variable, strictly speaking. I did this because samples were not refrigerated while I was processing them and the scanning was the bottleneck step, so samples later on were sitting on the counter longer.
```{r}
jgy_path <- "/Volumes/as_rsch_orianslab_tea01$/Image Analysis/2018/Leaf Scans/Manipulative/Jin Guan Yin"
lj_path <- "/Volumes/as_rsch_orianslab_tea01$/Image Analysis/2018/Leaf Scans/Manipulative/Longjing"

jgy_info <-
  dir(jgy_path, "*.jpg", full.names = TRUE) %>%
  file.info() %>% 
  as.tibble(rownames = "paths") %>% 
  mutate(plant_num = str_extract(paths, "\\d+(?=\\w?\\.jpg$)")) %>% 
  mutate(sample = paste0("J", plant_num)) %>% 
  mutate(mtime = if_else(sample == "J18",
                         ymd_hms("2018-07-30 23:17:00") + hours(4),
                         mtime)) %>% 
  mutate(time = mtime - min(mtime))

lj_info <-
  dir(lj_path, "*.png", full.names = TRUE) %>% 
  file.info() %>% 
  as.tibble(rownames = "paths") %>% 
  mutate(plant_num = str_extract(paths, "\\d+(?=\\w?\\.png$)")) %>% 
  mutate(sample = paste0("L", plant_num)) %>% 
  mutate(time = mtime - min(mtime))

processing_time <-
  bind_rows(jgy_info, lj_info) %>% 
  select(sample, plant_num, time) %>% 
  group_by(sample) %>% 
  summarize(time = min(time))


```

```{r}
lc2018_scaled <- 
  lc2018_2 %>% 
  mutate_at(vars(Caffeine:Theobromine), scale) %>% 
  mutate(cultivar = as.factor(cultivar)) %>% 
  left_join(processing_time)

rda.2018 <- 
  rda(lc2018_scaled %>% select(Caffeine:Theobromine) ~ mean_percent_damage*cultivar + Condition(as.numeric(time)),
      data = lc2018_scaled)
```

2. How much total variance does the experimental design explain?

```{r}
MVA.synt(rda.2018)
```

Looks like sample order doesn't matter. Only 1.86% of variation.

31.63% of variation explained by cultivar and leaf damage!

3. Test for significance that this amount of explained variance is higher than the null hypothesis of no effect of the experimental design:

```{r}
anova(rda.2018)
```

Highly significant global effect

4. Then we test for significance of individual factors and interaction terms:

```{r}
MVA.anova(rda.2018)
```

Significant effect of cultivar, no effect of damage.

5. Loadings

```{r}
rda.2018$CCA$v %>% 
  as_tibble(rownames = "compound") %>% 
  arrange(desc(abs(RDA1)))


MVA.plot(rda.2018, 
         fac = lc2018_scaled$cultivar,
         drawextaxes = FALSE,
         main = "LCMS compounds",
         main.pos = "topleft")
```

Cultivar differences mainly in Theobromine, Gallocatechin, Epigallocatechin and Caffeine with all of them lower in Jin Guan Yin