---
title: "2017 Total Polyphenols"
output: 
  html_notebook: 
    number_sections: yes
    toc: yes
    toc_float: yes
    highlight: kate
    theme: yeti
---

Total polyphenols were measured by Amma Agyei using the Folin-Ciocalteau assay Spring of 2019.

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(here)
library(car)
library(janitor)
library(cowplot)
library(broom)
library(glue)
library(chngpt)
```

# Read in data

Total polyphenols data, leafhopper density data, and leaf damage data

```{r message=FALSE, warning=FALSE}
tp2017 <- read_csv(here("data", "raw", "2017 total polyphenols.csv"))
damage2017 <- read_rds(here('data', 'cleaned', '2017_treatment_data.rds'))

tp2018 <- read_csv(here("data", "raw", "2018 total polyphenols.csv"))
damage2018 <- read_rds(here('data', 'cleaned', '2018_treatment_data.rds'))
```

Join data sheets and remove outliers identified previously.

```{r}
tp2017 <- 
  tp2017 %>% 
  clean_names() %>% 
  full_join(damage2017, by = "sample") %>%
  filter(sample != "Q2") %>% 
  select(-leaf_density) %>% #identical to `density_end`
  mutate(log_damage = log(mean_percent_damage))
tp2017
```
```{r}
tp2018 <- 
  tp2018 %>% 
  clean_names() %>% 
  full_join(damage2018, by = "sample") %>%
  filter(sample != "J23") %>% 
  mutate(log_damage = log(mean_percent_damage))
tp2018
```


- `sample`: cultivar and sample ID. Q = qing xin da mao
- `standard_deviation`: from averaging triplicates
- `mean_ga_equivalent`: galic acid equivalents in mg/mL (i think)
- `coefficient_of_variation`: from averaging triplicates(?)
- `leaf_density`: leafhopper density at end of experiment in insects/young leaf
- `mg_tea`: mg of tea powder
- `mg_g_gae`: galic acid equivalents in mg/g of tea powder


These data are means of triplicate runs.

# 2017

Total polyphenol data seem normal enough to use a regular linear model

```{r}
shapiro.test(tp2017$mg_g_gae)
#normal enough
```

## Polyphenols vs. Density

I'm going to apply the same models as I do for LCMS and GCMS compounds: null, linear, step, and hinge.  Then select the best fit by AIC

```{r}
dens.linear.17 <- lm(mg_g_gae ~ density_end, data = tp2017)
dens.null.17 <- lm(mg_g_gae ~ 1, data = tp2017)
dens.step.17 <- chngptm(mg_g_gae ~ 1, ~ density_end, type = "step", family = "gaussian", data = tp2017,
                        var.type = "bootstrap", ci.bootstrap.size = 100)
dens.hinge.17 <- chngptm(mg_g_gae ~ 1, ~ density_end, type = "hinge", family = "gaussian", data = tp2017)

bbmle::AICtab(dens.linear.17, dens.null.17, dens.step.17$best.fit, dens.hinge.17$best.fit)
summary(dens.step.17)
summary(dens.hinge.17)
```

Step model wins, but change in intercept isn't significantly different from zero.  Next is hinge, but slope isn't significant.  Null is next best.

```{r eval=FALSE, include=FALSE}
ggplot(augment(dens.step.17$best.fit), aes(.resid)) + geom_histogram(bins = 15)
ggplot(augment(dens.step.17$best.fit), aes(sample = .resid)) + geom_qq() + geom_qq_line()
```

```{r echo=FALSE}
dens.2017.plot <-
  ggplot(tp2017, aes(x = density_end, y = mg_g_gae)) +
  geom_point(alpha = 0.75, size = 2) +
  # geom_segment(aes(x = min(density_end),
  #                  xend = dens.step.17$chngpt,
  #                  y = dens.step.17$coefficients[1],
  #                  yend = dens.step.17$coefficients[1])) +
  # geom_segment(aes(x = dens.step.17$chngpt,
  #                  xend = max(density_end),
  #                  y = dens.step.17$coefficients[1] + dens.step.17$coefficients[2],
  #                  yend = dens.step.17$coefficients[1] + dens.step.17$coefficients[2])) +
  labs(x = "Final leahopper density (insects / young leaf)",
       y = "Total polyphenols (mg/g GAE)",
       # caption = "Sample Q2 removed",
       title = "2017") +
  theme_bw() 
dens.2017.plot
```

## Polyphenols vs. Damage

```{r}
# dam.poly.17 <- lm(mg_g_gae ~ poly(mean_percent_damage, 2), data = tp2017)
dam.linear.17 <- lm(mg_g_gae ~ log_damage, data = tp2017)
dam.null.17 <- lm(mg_g_gae ~ 1, data = tp2017)
dam.step.17 <- chngptm(mg_g_gae ~ 1, ~ log_damage, type = "step", family = "gaussian", data = tp2017,
                        var.type = "bootstrap", ci.bootstrap.size = 100, boot.test.inv.ci = TRUE)
dam.hinge.17 <- chngptm(mg_g_gae ~ 1, ~ log_damage, type = "hinge", family = "gaussian", data = tp2017,
                        var.type = "bootstrap", m.out.of.n = TRUE)

bbmle::AICtab(dam.linear.17, dam.null.17, dam.step.17$best.fit, dam.hinge.17$best.fit)
summary(dam.hinge.17)

summary(dam.step.17)


summary(dam.linear.17)
coefficients(dam.hinge.17)
# dam.hinge.17$coefficients[2]*dam.hinge.17$coefficients[1]
```
Now, the hinge model is the best, but the slope is not significant, same for step.  Linear is sigificant though


```{r echo=FALSE}
dam.2017.plot <-
  ggplot(tp2017, aes(x = log_damage, y = mg_g_gae)) +
  geom_point(alpha = 0.75, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size = 0.5) +
  # #hinge
  # geom_segment(aes(x = min(log_damage),
  #                  xend = dam.hinge.17$chngpt,
  #                  y = dam.hinge.17$coefficients[1],
  #                  yend = dam.hinge.17$coefficients[1])) +

  # geom_segment(aes(x = dam.hinge.17$chngpt,
  #                  xend = max(log_damage),
  #                  y = dam.hinge.17$coefficients[1],
                   # yend = dam.hinge.17$coefficients[1] + dam.hinge.17$coefficients[2])) +
  labs(x = "ln(mean % damaged leaf area)",
       y = "Total polyphenols (mg/g GAE)",
       title = "2017") +
  theme_bw() #+
dam.2017.plot
```


# 2018


## Regression with leafhopper density as predictor

Data seem normal enough to use a regular linear model

```{r}
shapiro.test(tp2018$mg_g_gae)
#normal enough
```


## Split cultivars
I've decided it's gonna be best to just analyze the two cultivars separately for everything
```{r}
tp_jgy <- tp2018 %>% filter(str_detect(sample, "J"))
tp_lj <- tp2018 %>% filter(str_detect(sample, "L"))
```

## Polyphenols vs. density

I'll test for a quadratic or linear relationship

### Jin Guan Yin
```{r}
# dens.jgy.poly <- lm(mg_g_gae ~ poly(density_end, 2), data = tp_jgy)
dens.jgy.lm <- lm(mg_g_gae ~ density_end, data = tp_jgy)
dens.jgy.null <- lm(mg_g_gae ~ 1, data = tp_jgy)
dens.jgy.step <- chngptm(mg_g_gae ~ 1, ~ density_end, type = "step", family = "gaussian", data = tp_jgy,
                        var.type = "bootstrap", ci.bootstrap.size = 100)
dens.jgy.hinge <- chngptm(mg_g_gae ~ 1, ~ density_end, type = "hinge", family = "gaussian", data = tp_jgy)
bbmle::AICtab(dens.jgy.lm, dens.jgy.null, dens.jgy.step$best.fit, dens.jgy.hinge$best.fit)
```

The null model wins. No effect of leafhopper density on total polyphenols for JGY

### Long Jing

```{r}
# dens.lj.poly <- lm(mg_g_gae ~ poly(density_end, 2), data = tp_lj)
dens.lj.lm <- lm(mg_g_gae ~ density_end, data = tp_lj)
dens.lj.null <- lm(mg_g_gae ~ 1, data = tp_lj)

dens.lj.step <- chngptm(mg_g_gae ~ 1, ~ density_end, type = "step", family = "gaussian", data = tp_lj,
                        var.type = "bootstrap", ci.bootstrap.size = 100)
dens.lj.hinge <- chngptm(mg_g_gae ~ 1, ~ density_end, type = "hinge", family = "gaussian", data = tp_lj)

bbmle::AICtab(dens.lj.lm, dens.lj.null, dens.lj.step$best.fit, dens.lj.hinge$best.fit)
summary(dens.lj.step)
summary(dens.lj.hinge)
summary(dens.lj.lm)
```

Step model wins, but change in intercept is not significant.  Same with hinge, same with linear, so null wins?

```{r}
#TODO add in step function line for longjing
dens.2018.plot <-
  ggplot(tp2018, aes(x = density_end, y = mg_g_gae, color = fct_relevel(cultivar, "L"), shape = fct_relevel(cultivar, "L"))) +
  geom_point(alpha = 0.75, size = 2) +
  # # step for longjing
  # geom_segment(aes(x = min(density_end),
  #                  xend = dens.lj.step$chngpt,
  #                  y = dens.lj.step$coefficients[1],
  #                  yend = dens.lj.step$coefficients[1]),
  #              color = "darkgreen") +
  # geom_segment(aes(x = dens.lj.step$chngpt,
  #                  xend = max(density_end),
  #                  y = dens.lj.step$coefficients[1] + dens.lj.step$coefficients[2],
  #                  yend = dens.lj.step$coefficients[1] + dens.lj.step$coefficients[2]),
  #              color = "darkgreen") +
  labs(x = "Final leafhopper density (insects / shoot)",
       y = "Total polyphenols (mg/g GAE)",
       title = "2018") +
  scale_color_manual(name = "Cultivar", values = c("J" = "purple", "L" = "darkgreen"),
                     labels = c("J" = "JGY", "L" = "LJ")) +
  scale_shape_discrete(name = "Cultivar", labels = c("J" = "JGY", "L" = "LJ")) +
  theme_bw() 
dens.2018.plot
```

## Polyphenols vs. damage
### Jin Guan Yin
```{r}
# dam.jgy.poly <- lm(mg_g_gae ~ poly(mean_percent_damage, 2), data = tp_jgy)
dam.jgy.lm <- lm(mg_g_gae ~ log_damage, data = tp_jgy)
dam.jgy.null <- lm(mg_g_gae ~ 1, data = tp_jgy)
dam.jgy.step <- chngptm(mg_g_gae ~ 1, ~ log_damage, type = "step", family = "gaussian", data = tp_jgy,
                        var.type = "bootstrap", ci.bootstrap.size = 100)
dam.jgy.hinge <- chngptm(mg_g_gae ~ 1, ~ log_damage, type = "hinge", family = "gaussian", data = tp_jgy)

bbmle::AICtab(dam.jgy.lm, dam.jgy.null, dam.jgy.step$best.fit, dam.jgy.hinge$best.fit)
```
Null model wins

### Long Jing
```{r}
# dam.lj.poly <- lm(mg_g_gae ~ poly(mean_percent_damage, 2), data = tp_lj)
dam.lj.lm <- lm(mg_g_gae ~ mean_percent_damage, data = tp_lj)
dam.lj.null <- lm(mg_g_gae ~ 1, data = tp_lj)
dam.lj.step <- chngptm(mg_g_gae ~ 1, ~ log_damage, type = "step", family = "gaussian", data = tp_lj,
                        var.type = "bootstrap", ci.bootstrap.size = 100)
dam.lj.hinge <- chngptm(mg_g_gae ~ 1, ~ log_damage, type = "hinge", family = "gaussian", data = tp_lj)
bbmle::AICtab(dam.lj.lm, dam.lj.null, dam.lj.step$best.fit, dam.lj.hinge$best.fit)
```

The winning model is the null model

```{r}
dam.2018.plot <-
  ggplot(tp2018, aes(x = log_damage, y = mg_g_gae, color = fct_relevel(cultivar, "L"), shape = fct_relevel(cultivar, "L"))) +
  geom_point(alpha = 0.75, size = 2) +
  labs(x = "ln(mean % damaged leaf area)",
       y = "Total polyphenols (mg/g GAE)",
       title = "2018") +
  scale_color_manual(name = "Cultivar", values = c("J" = "purple", "L" = "darkgreen"),
                     labels = c("J" = "JGY", "L" = "LJ")) +
  scale_shape_discrete(name = "Cultivar", labels = c("J" = "JGY", "L" = "LJ")) +
  theme_bw() 
dam.2018.plot
```

# Figures


## Density/damage vs. polyphenols
for both years, so 4 panel plot


```{r}
dam_dens_tp <-
  plot_grid(dam.2017.plot,
            dam.2018.plot +
              # ylim(3.8,8.7) +
              theme(legend.position = "none"),
            dens.2017.plot +
              labs(title = NULL),
            dens.2018.plot +
              labs(title = NULL) +
              theme(legend.position = "none"), 
            labels = "AUTO") %>% 
  plot_grid(get_legend(dam.2018.plot), rel_widths = c(1, 0.1))

dam_dens_tp
save_plot(here('figs', 'tp.png'), dam_dens_tp, ncol = 2, nrow = 2, base_asp = 1.1)
save_plot(here('figs', 'figure4.jpg'), dam_dens_tp, ncol = 2, nrow = 2, base_asp = 1.1)

```




