---
title: "2018 Leaf Scan Analysis"
output: html_notebook
---

Date initiated: 2019-05-06
Date compiled: `r Sys.Date()`
```{r}
library(here)
library(tidyverse)
library(glue)
```

# Choose random images for training set

First, copy a random subset of files to use as a training set.

```{r}
scandir <- "/Users/scottericr/Box/Image Analysis/2018/Leaf Scans/Manipulative"
leaves <- list.files(scandir, recursive = TRUE, pattern = "leaf.+\\.png", full.names = FALSE)
length(leaves)



df <- tibble(path = leaves)
df <- df %>%
  mutate(cultivar = str_extract(path, "^[^/]+(?=/)"),
         scan = str_extract(path, "(?<=/).+(?=/)"),
         leaf = str_extract(path, "[^/]+(?=.png)")) %>% 
  mutate(new.filename = glue("{scan}_{leaf}.png"),
         from.path = glue("{scandir}/{path}"),
         to.path = glue("{scandir}/Training Sample/{new.filename}"))

# runif(1, 1, 1000)
set.seed(688)
train <- df %>% group_by(cultivar) %>% sample_n(15) %>% ungroup()
# file.copy(from = train$from.path, to = train$to.path)
```

Output a list of file paths. This can be used by FIJI in file > import > Stack From List... to make a virtual stack. However, I think it crops and/or resizes leaves, which isn't optimal.

```{r}
# write_csv(select(train, to.path), glue("{scandir}/Training Sample/file_list.csv"),col_names = FALSE)
list.files(glue("{scandir}/Training Sample/"), pattern = "*.png", full.names = TRUE) %>% 
  write_lines(glue("{scandir}/Training Sample/file_list.csv"))
```

Instead, I should open all the images and convert to stack manually.  I've previously used this macro to automatically make stacks.  I should edit it to make stacks of ~30 for later.

```{javascript}
dir = getDirectory("choose a folder");

list = getFileList(dir);
nstacks = -floor(-list.length/5);
print(nstacks);
stacksize = round(list.length/nstacks);

//for each stack:
for(n=0; n<nstacks; n++){
	//open the first set of files
	files = Array.slice(list, stacksize*(n), stacksize*(n+1));
	for (i=0; i<files.length; i++){
			open(files[i]);
		}
	run("Images to Stack", "method=[Copy (top-left)] name=Stack title=[] use");
	width = getWidth;
	height = getHeight;
	setColor("white");
	for (j=1; j<=nSlices; j++){
		setSlice(j);
		floodFill(width-1, height-1);
	}

	saveAs("tiff", dir + "/Stack_" + n + ".tif");
	close();
}
```

